---
title: "ProFound: Beginner's Guide"
author: "ProFound Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ProFound: Beginner's Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.width = 7,
  fig.height = 7
)
```

## What is ProFound?

ProFound is an R package designed to **automatically detect and extract astronomical objects from images**. If you have an image (like a telescope observation), ProFound can:

- Find all the stars, galaxies, or other objects in the image
- Determine their locations and sizes
- Measure their brightness
- Separate overlapping objects
- Create a catalogue you can analyse further

Think of it as a "find all the things in this picture" tool, with sophisticated algorithms to handle real astronomical data.

## When Would You Use ProFound?

Typical use cases:

- **Photometry**: Measuring the brightness of stars or galaxies
- **Source catalogues**: Creating a list of all objects detected in a survey
- **Image analysis**: Understanding what's in your astronomical images
- **Pre-processing**: Preparing data before further analysis (like spectroscopy)

## The Main Function: `profoundProFound()`

The core of ProFound is the `profoundProFound()` function. Most of the time, this is the only function you need to know about.

### Basic Usage

The simplest form requires just an image:

```{r eval=FALSE}
library(ProFound)

# Run ProFound on an image
result <- profoundProFound(image = my_image)
```

That's it! ProFound will:
1. Estimate the background sky level
2. Find connected regions above a threshold
3. Clean up the detections
4. Measure properties of each object
5. Return a list with all the results

### What Do You Need to Input?

**The image:**
- Must be a 2D matrix or array of pixel values
- Typically read from a FITS file (use `Rfits_read_image()` from the `Rfits` package)
- Can be from any telescope or camera

```{r eval=FALSE}
library(Rfits)

# Read a FITS file
fits_data <- Rfits_read_image("my_observation.fits")

# Run ProFound
result <- profoundProFound(image = image)
```

### Key Output: The Catalog

The most useful output is `result$segstats`, a data frame containing detected objects and various useful properties, including:

- `segID`: Unique ID for each object
- `x`, `y`: Centre position (in pixels)
- `N100`: Number of pixels in the object
- `flux`: Total brightness (in ADU or counts)
- `mag`: Magnitude (if `magzero` is set properly this will be in AB mag)
- `R50`: Size of the object (radius containg 50% of the flux)

```{r eval=FALSE}
# View the catalog
head(result$segstats)

# Access specific objects
bright_objects = result$cat[result$cat$segstats$flux > 1000, ]
```

## Common Parameters Explained

ProFound has many parameters, but most have sensible defaults. Here are the ones you might adjust:

### `skycut` and `pixcut`

These control how aggressively ProFound detects objects:

- **`skycut`** (default: 1.5): Detected objects must be this many times the sky noise above the background
- **`pixcut`** (default: 3): Minimum number of connected pixels to count as an object

*Lower values* = more detections (including fainter objects and noise)  
*Higher values* = fewer detections (only brightest objects)

```{r eval=FALSE}
# Detect fainter objects
result_faint <- profoundProFound(image, skycut = 1, pixcut = 2)

# Detect only bright objects
result_bright <- profoundProFound(image, skycut = 3, pixcut = 5)
```

### `magzero`

By default, ProFound reports brightness in raw counts (ADU). To convert to magnitudes:

```{r eval=FALSE}
# Tell ProFound the AB magnitude of a reference zero point
result <- profoundProFound(image, magzero = 25)  # Example: 25 mag zero point
# Now result$cat$segstats will be in AB magnitudes
```

### `mask`

If certain parts of your image are bad (cosmic rays, detector artifacts), provide a mask:

```{r eval=FALSE}
# Create a mask: TRUE for good pixels, FALSE for bad pixels
bad_pixels <- image$imDat < -100  # Example: very negative pixels are bad

result <- profoundProFound(image, mask = bad_pixels)
# ProFound will ignore bad pixels
```

### `gain` and `pixscale`

For more accurate measurements:

```{r eval=FALSE}
# Tell ProFound the AB magnitude of a reference zero point
# gain: electrons per ADU (from your detector specs)
# pixscale: arcseconds per pixel (from your telescope/camera)
result <- profoundProFound(image, magzero = 25, gain = 2.5, pixscale = 0.5)
```

## A Complete Example

```{r}
library(ProFound)
library(Rfits)

# 1. Read your image
image = Rfits_read_image(system.file("extdata", 'VIKING/mystery_VIKING_Z.fits',
  package="ProFound"))

# 2. Create a simple mask (remove obvious bad pixels)
mask <- image$imDat < -10  # Ignore very negative pixels

# 3. Run ProFound with sensible settings for detecting galaxies
result <- profoundProFound(
  image = image,
  mask = mask,
  skycut = 1.5,      # Moderate sensitivity
  pixcut = 3,        # Require at least 3 connected pixels
  magzero = 26,      # Convert to magnitudes
  pixscale = 0.3,    # Pixel scale in arcsec
  plot = TRUE        # Show diagnostic plots
)

# 4. Examine the results
cat_objects <- result$segstats

# How many objects did we detect?
nrow(cat_objects)

# What's the brightest object? (Usually the first object with segID=1, but not always)
brightest <- cat_objects[which.max(cat_objects$flux), ]

# Plot a histogram of magnitudes
hist(cat_objects$mag, breaks = 50, 
     main = "Magnitude Distribution",
     xlab = "Magnitude")
```

Often the only bit people want to save is the *segstats* catalogue:

```{r eval=FALSE}
# Export the catalog
write.csv(cat_objects, "/path/to/use/detected_objects.csv", row.names = FALSE)
```

## Understanding the Segmentation Image

ProFound also returns `result$segim`, a "segmentation image":

- Each pixel is labelled with the ID of the object it belongs to
- Value 0 = background (no object)
- Value 1, 2, 3, ... = different detected objects

```{r}
# How many unique objects were detected?
length(unique(as.integer(result$segim))) - 1L

# Which will be the same as
dim(result$segstats)

# Visualize it
magimage(result$segim, col = rainbow(max(result$segim)))
```

## Troubleshooting

### Too many detections (too much noise)?

Increase `skycut` or `pixcut`:

```{r}
profoundProFound(image, skycut = 2.5, pixcut = 5, plot = TRUE)
```

### Missing faint objects?

Decrease `skycut` or `pixcut`:

```{r}
profoundProFound(image, skycut = 0.5, pixcut = 2, plot = TRUE)
```

### Objects merged together?

Lower de-blending tolerance:

```{r}
profoundProFound(image, tolerance = 1, plot = TRUE)
```

### Wrong background estimate?

Try a different sky estimation method:

```{r}
profoundProFound(image, skytype = "mean", plot = TRUE)
```

## Next Steps

- Check `?profoundProFound` for the full parameter list
- Look at other vignettes for advanced topics (de-blending, photometry, etc.)
- Run `profoundProFound(..., plot = TRUE)` to see diagnostic plots
- Get advanced plots with plot method run on `profoundProFound` output, i.e. `plot(result)`

Good luck with your analysis!
